Phylogenetics and the human microbiome.

### 2014-07-18 sbreview.tex
Formatting.
Thank peer reviewers
Arne's version of the unifrac sentence.

\newcommand{\pdLegend}{
    Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as circles and abundances are shown as the size of the circles.
    Unweighted PD takes the total length of branches sitting between tree tips represented in a sample.
    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the total area of the edges.
    One such abundance-weighted measure simply takes the absolute value of the difference of the total read abundance on one side compared to the other.
}
\forarxiv{
\begin{wrapfigure}[26]{r}{2.2in}
  % \capstart
  \vspace{-25pt}
  \begin{center}
    \includegraphics[width=2.2in]{figures/pd.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\pdLegend}
  \label{fig:pd}
\end{wrapfigure}
}

@@

To calculate the UniFrac divergence between two samples of reads, one builds a tree on both samples and then performs a calculation on the branch lengths according whether branches have descendants in both samples or one sample only.
Specifically, the original UniFrac divergence is the fraction of the total branch length of the tree coming from branches with descendant leaves in exactly one of the two samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
The earth-movers distance \citep{monge1781memoire,Villani2003-wv} between two probability distributions in this case can be defined using a physical analogy as the minimum amount of ``work'', defined as mass times distance, required to move probability mass in one distribution to another along the tree.
In this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
There are clear connections between UniFrac and phylogenetic diversity explored by \citet{faith2009cladistic}, who point out related measures in Faith's earlier work.

@@

\section{Acknowledgements}
I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 MCEB Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank Bastien Boussau, Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript.
This work was supported by National Institutes of health grant R01-HG005966-01 and National Science Foundation grant 1223057.

>>>>

\newcommand{\pdLegend}{
    Unweighted phylogenetic diversity (PD, left) and an abundance-weighted PD measure (right), where taxa present in a sample are shown as circles and abundances are shown as the size of the circles.
    Unweighted PD takes the total length of branches sitting between tree tips represented in a sample.
    Abundance-weighted measures take a weighted sum of branch lengths where weight is determined in some way by the abundance of the taxa on either side of the branch: if we give edges width according to their weight, the abundance-weighted measure can be thought of as the sum of the total area of the edges.
    One such abundance-weighted measure simply takes the absolute value of the difference of the total read abundance on one side compared to the other.
}
\forarxiv{
\begin{wrapfigure}[27]{r}{2.2in}
  % \capstart
  % \vspace{-25pt}
  \begin{center}
    \includegraphics[width=2.2in]{figures/pd.pdf}
  \end{center}
  \vspace{-10pt}
  \caption{\pdLegend}
  \label{fig:pd}
\end{wrapfigure}
}

@@

To calculate the UniFrac (``unique fraction'') divergence between two samples of reads, one builds a global tree based on both samples and then calculates the fraction of the total tree length found in only one (i.e. the fraction of the tree unique to one) of the two samples.
Specifically, the original UniFrac divergence is the fraction of the total branch length of the tree coming from branches with descendant leaves in exactly one of the two samples \citep[Fig.~\ref{fig:unifrac};][]{LozuponeKnightUniFrac05}.
Weighted UniFrac is an abundance weighted version \citep{LozuponeEaWeightedUnifrac07}.
These dissimilarity measures have hundreds of citations.
\citet{evans2012phylogenetic} showed that weighted UniFrac is in fact a specific case of the earth-movers distance, and that the commonly-used randomization procedure for significance estimation has a central limit theorem approximation.
The earth-movers distance \citep{monge1781memoire,Villani2003-wv} between two probability distributions in this case can be defined using a physical analogy as the minimum amount of ``work'', defined as mass times distance, required to move probability mass in one distribution to another along the tree.
In this case the size of the dirt piles is proportional to the number of reads mapping to that location in the tree (Fig.~\ref{fig:dirtpiles}).
\citet{chen2012associating} have shown that a partially abundance-weighted variant of UniFrac may have greater power to resolve community differences than either unweighted or weighted UniFrac.
There are clear connections between UniFrac and phylogenetic diversity explored by \citet{faith2009cladistic}, who point out related measures in Faith's earlier work.

@@

\section{Acknowledgements}
I thank Olivier Gascuel for the opportunity to present on this subject during the 2013 MCEB Mathematical and Computational Evolutionary Biology workshop and for organizing a corresponding special section of \textit{Systematic Biology}.
I am grateful to Aaron Darling, David Fredricks, Noah Hoffman, Steven Kembel, Connor McCoy, Martin Morgan, and Sujatha Srinivasan for interesting discussions that informed this review, thank Bastien Boussau, Noah Hoffman, Christopher Small and Bj\"orn Winckler for providing feedback on the manuscript.
The manuscript was greatly improved by thoughtful peer review from Frank (Andy) Anderson, Olivier Gascuel, Alexis Stamatakis, Frédéric Delsuc, and an anonymous referee.
This work was supported by National Institutes of health grant R01-HG005966-01 and National Science Foundation grant 1223057.
